---
title: "dtrackr - Consort statement example"
output: rmarkdown::html_vignette
# output: html_document
# output: pdf_document
# output: word_document
vignette: >
  %\VignetteIndexEntry{dtrackr - Consort statement example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: dtrackr.bib
---

# CONSORT statement

CONSORT diagrams are part of the requirements in reporting parallel group clinical trials or case control designs in observational studies. They are described in the updated 2010 CONSORT statement [@schulzCONSORT2010Statement2010]. They clarify how patients were recruited, selected, randomised and followed up. For observational studies an equivalent requirement is the STROBE statement [@vonelmStrengtheningReportingObservational2008]. There are other similar requirements for other types of study such as the TRIPOD statement that are applicable for multivariate models [@collinsTransparentReportingMultivariable2015].

For this demonstration we use the `cgd` data set from the `survival` package [@survival-book; @survival-package], which is from a placebo controlled trial of gamma interferon in Chronic Granulomatous Disease.


```{r include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

here::i_am("vignettes/consort-example.Rmd")

library(tidyverse)
library(dtrackr)
```

In this example the `treatment` column contains the intervention. 

For the analysis we are considering only the first observations for each patient, and
our study criteria are as follows:

* Only patients from US based centres are considered
* Patients must not be on steroids at inception.
* Patients must be aged between the ages of 5 and 35 inclusive.

This can be coded into the `dplyr` pipeline, with additional `dtrackr` functions:

```{r}
# Some useful formatting options
old = options(
  dtrackr.strata_glue="{tolower(.value)}",
  dtrackr.strata_sep=", ",
  dtrackr.default_message = "{.count} records",
  dtrackr.default_headline = NULL
)

demo_data = survival::cgd %>% 
  track() %>%
  filter(enum == 1, .type="inclusion", .messages="{.count.out} first observation") %>%
  include_any(
    hos.cat == "US:NIH" ~ "{.included} NIH patients",
    hos.cat == "US:other" ~ "{.included} other US patients"
  ) %>%
  group_by(treat, .messages="cases versus controls") %>%
  comment() %>%
  capture_exclusions() %>%
  exclude_all(
    age<5 ~ "{.excluded} subjects under 5",
    age>35 ~ "{.excluded} subjects over 35",
    steroids == 1 ~ "{.excluded} on steroids at admission"
  ) %>%
  comment(.messages = "{.count} after exclusions") %>%
  status(
    mean_height = sprintf("%1.2f \u00B1 %1.2f",mean(height),sd(height)),
    mean_weight = sprintf("%1.2f \u00B1 %1.2f",mean(weight),sd(weight)),
    .messages = c(
      "average height: {mean_height}",
      "average weight: {mean_weight}"
    )                    
  ) %>%
  ungroup(.messages = "{.count} in final data set")

# restore to originals
options(old)
```

With a bit of experimentation the flowchart needed for a STROBE/CONSORT checklist can be generated. One option to output the flowchart is `svg` which can then be manually formatted as required, but for publication ready output `pdf` is usually preferred.

```{r include=FALSE}
# saving this flowchart for the JOSS paper.
if (interactive()) demo_data %>% flowchart(filename = here::here("vignettes/joss/figure1-consort.pdf")) %>% invisible()
```

```{r}
demo_data %>% flowchart()
```

# Excluded data

During this pipeline, we may be keen to understand why certain data items are being rejected. This would enable us to examine the source data, and potentially correct it during the data collection process. We've used it to allow continuous quality checks on the data to feed back to the data curators, as we regularly conduct analyses. By tracking the exclusions, not only do we track the data flow through the pipeline we also retain all excluded items, with the reason for exclusion. Thus we can reassure ourselves that the exclusions are as expected. We enabled this by calling `capture_exclusions()` in the pipeline above. Having tracked the exclusions we can retrieve them by calling `excluded()` which gives a data frame with the excluded records and the reasons. If the exclusions happened over multiple stages as the dataframe format change in between then this will be held as a nested dataframe (i.e. see `?tidyr::nest`):

```{r}

# here we filter out the majority of the actual content of the excluded data to focus on the 
# metadata recovered during the exclusion.
demo_data %>% excluded() %>% select(.stage,.message,.filter,age, steroids)

```

This list may have multiple entries for a single data item, if for example something is excluded in any one step for many reasons.

# Tagging the pipeline

For reporting results it is useful to have the numbers from the flowchart to embed into the text of the results section of the write up. Here we show the same pipeline as above, but with 4 uses of the `.tag` system for labelling part of the pipeline. This captures data in a tag-value list during the pipeline, and retains it as metadata for later reuse. 

```{r}

demo_data = survival::cgd %>% 
  track(.messages = NULL) %>%
  filter(enum == 1, .type="inclusion", .messages="{.count.out} first observation") %>%
  comment(.tag = "initial cohort") %>%
  #         ^^^^^^^^^^^^^^^^^^^^^
  #         TAGS DEFINED
  
  include_any(
    hos.cat == "US:NIH" ~ "{.included} NIH patients",
    hos.cat == "US:other" ~ "{.included} other US patients"
  ) %>%
  group_by(treat, .messages="cases versus controls") %>%

  comment(.tag="study cohort") %>%
  #       ^^^^^^^^^^^^^^^^^^^
  #       SECOND SET OF TAGS DEFINED
  
  capture_exclusions() %>%
  exclude_all(
    age<5 ~ "{.excluded} subjects under 5",
    age>35 ~ "{.excluded} subjects over 35",
    steroids == 1 ~ "{.excluded} on steroids at admission"
  ) %>%
  
  comment(.messages = "{.count} after exclusions") %>%
  
  status(
    mean_height = sprintf("%1.2f \u00B1 %1.2f",mean(height),sd(height)),
    mean_weight = sprintf("%1.2f \u00B1 %1.2f",mean(weight),sd(weight)),
    .messages = c(
      "average height: {mean_height}",
      "average weight: {mean_weight}"
    ),
    .tag = "qualifying patients"
  #  ^^^^^^^^^^^^^^^^^^^^^^^^^^^^
  #  THIRD SET TAGS DEFINED                    
  ) %>%
  ungroup(.messages = "{.count} in final data set", .tag="final set")
  #                                                 ^^^^^^^^^^^^^^^^
  #                                                 LAST TAGS DEFINED

```

The tagged data can be retrieved as follows, which will give you all tagged data for all 4 points in the pipeline:

```{r}
demo_data %>% tagged() %>% tidyr::unnest(.content)
```

More often though you will want to retrieve specific values from specific points for the results text for example:

```{r}
initialSet = demo_data %>% tagged(.tag = "initial cohort", .glue = "{.count} patients")
finalSet = demo_data %>% tagged(.tag = "final set", .glue = "{.count} patients")

# there were `r initialSet` in the study, of whom `r finalSet` met the eligibility criteria.
```

For example there were `r initialSet` in the study, of whom `r finalSet` met the eligibility criteria.

More complex formatting and calculations are made possible by use of the `glue` specification, including those that happen on a per group basis, and we can also pull in values from elsewhere in our analysis.

```{r}
demo_data %>% tagged(
    .tag = "qualifying patients", 
    .glue = "{.strata}: {.count}/{.total} ({sprintf('%1.1f', .count/.total*100)}%) patients on {sysDate}, with a mean height of {mean_height}", 
    sysDate = Sys.Date()
    # we could have included any number of other parameters here from the global environment
  ) %>% dplyr::pull(.label)
```

Sometimes it will be necessary to operate on all tagged content at once. This is
possible but be aware that the content available depends somewhat on where the
tag was set in the pipeline so not all fields will always be present (although
`.count` and `.total` will be). The `.total` is the overall number of cases at
that point in the pipeline. `.count` is the number of cases in each strata.

```{r}
demo_data %>% tagged(.glue = "{.count}/{.total} patients")
```

For comparing inclusions and exclusions at different stages in the pipeline
using tags the following example may be useful:

```{r}
demo_data %>% 
  tagged() %>%   # selects only top level content
  tidyr::unnest(.content) %>% 
  dplyr::select(.tag, .total) %>% 
  dplyr::distinct() %>%
  tidyr::pivot_wider(values_from=.total, names_from=.tag) %>% 
  glue::glue_data("Out of {`initial cohort`} patients, {`study cohort`} were eligible for inclusion on the basis of their age
  but {`study cohort`-`qualifying patients`} were outside the age limits. 
  This left {`final set`} patients included in the final study (i.e. overall {`initial cohort`-`final set`} were removed).")
```

# Reusable functions 

Composing `dtrackr` inclusion and exclusion criteria into functions lets you
reuse them across different studies, but to be useful these functions need to be
parameterised. As `dtrackr` already uses formulae as part of the input to
construct the formulae for `dtrackr` from a helper function requires injection
support, using `rlang::inject()`. This is an advanced R topic but the following
example gives a sense of what is possible.

```{r}

# This is a reusable function to restrict ages
age_restrict = function(df, age_col, min_age = 18, max_age = 65) {
  age_col = rlang::ensym(age_col)
  message = sprintf("{.included} between\n%d and %d years", min_age, max_age)
  dtrackr::include_any(df,
    # injection support for parameters must be made explicit using
    # rlang::inject in any functions using include_any or exclude_all
    rlang::inject(min_age <= !!age_col & max_age >= !!age_col ~ !!message)
  )
}

survival::cgd %>% 
  # the `age` column is in 
  age_restrict(age, max_age = 30) %>%
  # demonstrating that this works in 2 stages
  age_restrict(age, min_age = 20) %>% 
  flowchart()

```

# References

<div id="refs"></div>
